Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-04-10 VAST_submission/vgtc_journal_latex/paper.tex

\subsubsection{April 11 to April 15, 2011}
%The next patch chronologically comes in 2011, from April 11 to April 15.   
On the last two days of %the 
this patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.   Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.  The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) where he is considering leaving Benghazi \cite{StevensUpdate}.

@@

\subsubsection{Two patches in December 2012}
The last two patches appear 
shortly after the email gap from Section~\ref{sect:emailGap}.  As seen in Figure~\ref{fig:ZeroEmailsDec2012}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ZeroEmailsDec2012}
\caption{Dec 3 -- Jan 14, 2012:  Selected Emails are from Clinton; Scheduled foreign travel marked in blue.}
\label{fig:ZeroEmailsDec2012}
\end{center}
\end{figure}
these are the two closely related gaps in December 2012 which,
% for the purposes of discourse only,  we are counting as only one of our 
%ten patches
might be counted as one large gap.
Similar to June 2009, there is an easy explanation at hand, at least for the first of these.  Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.  She had `already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.

@@

As seen in Section~\ref{sect:youHaveNoMail} this was often sufficient to turn up information that explained the lack of email as naturally occurring due to injury or travel.  In other cases, the information turned up was possibly contentious and could be interpreted to suggest that the email was deliberately missing.  We do not, and cannot, suggest that this is the case.  Of course, the way in which the emails were selected to be handed over to the State Department and the subsequent erasure of all back up copies does little to dissuade anyone who might hold this view.

We were surprised to see how often something contentious showed up with the Google search for the identified patches.   To check whether such a search would nearly always turn up something contentious, we randomly selected 10 combinations of months and years over the time period of the archive (a total of 60 possible combinations).  One of the combinations we turned up matched a patch we had identified ("Dec 2012"), the remaining did not.  The complete sample (and what we found for each) was as follows:  ``June 2011''  (pro forma cable from SECSTATE that State Dept. employees change their passwords), ``Nov 2010''  (U.S. Embassy Cables appear on Wikileaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on GlobalFood Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J. Crowley forced to resign), ``May 2012'' (State Dept. investigating computer shipments to Iran), ``Dec 2011'' (Putin accuses Clinton of encouraging Russian protests), ``June 2012'' (Geneva I conference on Syria; Bill Clinton checked with State Dept. on paid speech to group with Tehran ties; State Dept. Trafficking in Persons report), and  ``Oct 2011'' (Iranian plot to kill Saudi ambassador, State Dept. cannot confirm Gaddafi's death).   None but ``Dec 2012'' have a connection to the Benghazi controversy though two mention email.

 We need to also point out that the extraction of data is always an opportunity for error to occur.   For example, we have already mentioned difficulties with processing email addresses that have been partially or entirely redacted.  The Wikileaks transcription from pdf to HTML is also imperfect.  For example, the string ``b9''  appears in one email record (Doc ID: 10257) in the Wikileaks transcription\cite{errorWikileaks} as a transcription error of the day ``09'' from the original pdf.  This might be grabbed as a B9  redaction code, if one is not careful.  The same email record also has ``To'' and ``From'' header which is empty except for a date.  This header does not appear in the original pdf; rather the pdf begins with the next one inside the HTML.   There are no doubt other errors as well.

@@

First, metadata is often more reliable, regular, and, being easily collected, more generally available.  For Secretary Clinton's emails,  some metadata is lost, either because their source is a printed form or because it has been redacted.  Examining metadata also arguably intrudes least upon the privacy of the individual correspondents, compared at least to the email's content.

>>>>

\subsubsection{April 11 to April 15, 2011}
%The next patch chronologically comes in 2011, from April 11 to April 15.   
On the last two days of %the 
this patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.   Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.  The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) and that Ambassador Stevens is considering leaving Benghazi \cite{StevensUpdate}.

@@

\subsubsection{Two patches in December 2012}
The last two patches appear 
shortly after the email gap from Section~\ref{sect:emailGap}.  As seen in Figure~\ref{fig:ZeroEmailsDec2012}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ZeroEmailsDec2012}
\caption{Dec 3 -- Jan 14, 2012:  Selected Emails are from Clinton; Scheduled foreign travel marked in blue.}
\label{fig:ZeroEmailsDec2012}
\end{center}
\end{figure}
these are the two closely related gaps in December 2012 which,
% for the purposes of discourse only,  we are counting as only one of our 
%ten patches
might be counted as one large gap.
Similar to June 2009, there is an easy explanation at hand, at least for the first of these.  Clinton had suffered a concussion ``\ldots in a fall brought on by an illness'' \cite{concussion} and was recovering at home by December 15.  She had ``already been forced to cancel a planned trip to the Middle East and north Africa'' \cite{concussion} that week; this is not the foreign travel shown in Figure~\ref{fig:ZeroEmailsDec2012} \cite{ForeignSched}.

@@

As seen in Section~\ref{sect:youHaveNoMail} this was often sufficient to turn up information that explained the lack of email as naturally occurring due to injury or travel.  In other cases, the information turned up was possibly contentious and could be interpreted to suggest that the email was deliberately missing.  We do not, and cannot, say that this is the case.  Of course, the way in which the emails were selected to be handed over to the State Department and the subsequent erasure of all back up copies does little to dissuade anyone who might hold this view.

We were surprised to see how often something contentious showed up with the Google search for the identified patches.   To check whether such a search would nearly always turn up something contentious, we randomly selected 10 combinations of months and years over the time period of the archive (a total of 60 possible combinations).  One of the combinations we turned up matched a patch we had identified ("Dec 2012"), the remaining did not.  The complete sample (and what we found for each) was as follows:  ``June 2011''  (SECSTATE cables State Dept. employees to change their passwords), ``Nov 2010''  (U.S. Embassy Cables appear on WikiLeaks; Clinton rejects suggestion by Huma Abedin to start using  a government email account),   ``Dec 2012'' (documented in Section~\ref{sect:youHaveNoMail}), ``March 2009'' (Clinton hits reset button with Russia), ``May 2011'' (Clinton delivers remarks on Global Food Security), ``March 2011'' (Clinton rejects talks with Gaddafi's sons; P.J. Crowley forced to resign), ``May 2012'' (State Dept. investigating computer shipments to Iran), ``Dec 2011'' (Putin accuses Clinton of encouraging Russian protests), ``June 2012'' (Geneva 1 conference on Syria; Bill Clinton checked with State Dept. on paid speech to group with Tehran ties; State Dept. Trafficking in Persons report), and  ``Oct 2011'' (Iranian plot to kill Saudi ambassador; State Dept. cannot confirm Gaddafi's death).   None but ``Dec 2012'' have a connection to the Benghazi controversy, though two mention email.

 We need to also point out that the extraction of data is always an opportunity for error to occur.   For example, we have already mentioned difficulties with processing email addresses that have been partially or entirely redacted.  The Wikileaks transcription from pdf to HTML is also imperfect.  For example, the string ``b9''  appears in one email record (Doc ID: 10257) in the Wikileaks transcription\cite{errorWikileaks} as a transcription error of the day ``09'' from the original pdf.  This might be grabbed as a B9  redaction code, if one is not careful.  The same email record also has ``To:'' and ``From:'' header which is empty except for a date.  This header does not appear in the original pdf; rather the pdf begins with the next one inside the HTML.   There are no doubt other errors as well.

@@

First, metadata are often more reliable, regular, and, being easily collected, more generally available.  For Secretary Clinton's emails,  some metadata is lost, either because their source is a printed form or because it has been redacted.  Examining metadata also arguably intrudes least upon the privacy of the individual correspondents, compared at least to the email's content.
